---
title: "AE03 - Exploring gerrymandring"
author: "Rafael Wenslaff"
date: "2026-09-10"
format: html
execute: 
  warning: false
  message: false
---

## Getting started

Load the `tidyverse`, `usdata`, and `sjPlot` packages, and get the `gerrymander` data out of the `usdata` package. View the data using `glimpse`.

```{r}
library(tidyverse)
library(usdata)
library(sjPlot)
library(gtsummary)

gerrymander <- usdata::gerrymander
```

## 1. Congressional districts per state

Which state has the most congressional districts? The least? How many congressional districts are there in this state? Use functions such as `table`, `tbl_summary` or `count`.

```{r}
gerrymander %>%
  count(state) %>%
  arrange(desc(n))
```

* CA has the most congressional districts.

```{r}
gerrymander %>%
  count(state) %>%
  arrange(n)
```

* AK, DE, MT, ND, SD, VT, and WY have the least congressional districts.

## 2. Districts at the tails

Make side-by-side box plots of percent of vote received by Trump in the 2016 Presidential Election by prevalence of gerrymandering. Identify any Congressional Districts that are potential outliers — points that fall well outside the whiskers of their box, meaning that district's Trump vote share was unusually high or low compared to other districts with the same gerrymandering prevalence. What information can you share about them?

```{r}
ggplot(gerrymander, aes(x = trump16, y = gerry)) +
  geom_boxplot() +
  labs(
    x = "Votes received (%)",
    y = "Level of gerrymandering",
    title = "Percentage of Trump votes in the 2016 Presidential Election",
    subtitle = "by gerrymandering prevalence"
  ) 
```
```{r}
gerrymander %>%
  filter(trump16 < 10, gerry == "low") %>%
  select(-state, -dem18, -gerry)
```
* The congressional districts that are potential outliers are NY-13 and NY-15. They had unusually low Trump vote shares, with Trump receiving 5.4% of NY-13 votes and 4.9% of NY-15 votes. Both districts had democratic election winners, Adriano Espaillat (NY-13) and Jose Serrano (NY-15). 

## 3. Gerrymandering and flipping

This relationship can be looked at from two directions — does gerrymandering predict a 2018 flip, and does a 2018 flip predict gerrymandering? Support **both** with a visualization and summary statistics. The plot answers "what does the mix look like within each group?"; the table gives the exact proportions the plot is showing. Write 1-2 bullet points about what you noticed from each plot. Use specific numbers in your answer.

**Direction 1:** Is a Congressional District more likely to have high prevalence of gerrymandering if a Democrat was able to flip the seat in the 2018 election? To compare the breakdown of gerrymandering prevalence within each flipped group, make sure `flip18` is on the x axis and that your percents add up to 100% within each category.

```{r}
ggplot(gerrymander, aes(x = flip18, fill = gerry)) +
  geom_bar(position = "fill") +
  labs(
    x = "2018 flip",
    y = "Proportion",
    title = "Gerrymandering prevalence by 2018 seat flip",
    subtitle = "proportion of districts within each flip category"
  )
```
```{r}
gerrymander %>%
  group_by(flip18) %>%
  summarise(
    low = 100 * sum(gerry == "low") / n(),
    mid = 100 * sum(gerry == "mid") / n(),
    high = 100 * sum(gerry == "high") / n()
  )
```
* This data suggests that Congressional Districts are more likely to have `high` levels of gerrymandering if the Democrat was unable to flip the seat. While `mid` levels of gerrymandering have the majority in both cases, flipped seats are more likely to correlate to `low` levels (21.05%) of gerrymandering over `high` levels (13.16%), while unflipped seats are more likely to correlate to `high` levels (25.00%) of gerrymandering over `low` levels (13.27%).

  (It is unclear what a value of -1 represents for `flip18`, as it is not explicitly mentioned in the documentation).

**Direction 2:** Is a Congressional District more likely to be flipped to a Democratic seat if it has high prevalence of gerrymandering, or low prevalence? Same two variables, just flipped — so now `gerry` is on the x-axis. Again, make sure your percents add up to 100% within each category.

```{r}
ggplot(gerrymander, aes(x = gerry, fill = factor(flip18))) +
  geom_bar(position = "fill") +
  labs(
    x = "Level of gerrymandering",
    y = "Proportion",
    fill = "flip18",
    title = "2018 seat flip by gerrymandering prevalence",
    subtitle = "Proportion of distrcts within each gerrymandering level"
  )
```

```{r}
gerrymander %>%
  group_by(gerry) %>%
  summarise(
    unknown = 100 * sum(flip18 == -1) / n(),
    no = 100 * sum(flip18 == 0) / n(),
    yes = 100 * sum(flip18 == 1) / n()
  )
```

* This data suggests that Congressional Districts are most likely to be flipped to a Democratic seat if they have a `low` prevalence of gerrymandering. `high` levels of gerrymandering correlate to 4.85% flipped seats, while `low` levels of gerrymandering correlate to 12.90% flipped seats. `mid` levels of gerrymandering follow the trend, with a 9.25% proportion of flipped seats, which is between the `high` and `low` proportions.

## 4. Gerrymandering and presidential margin

Using `mutate`, create a new variable, `margin16`, measuring how many percentage points Clinton won or lost by in each district (`clinton16 - trump16`). Does gerrymandering prevalence relate to how lopsided the 2016 presidential race was in that district? Report the min, mean, and max margin by `gerry` level, then visualize the full distribution with a violin + boxplot combo. Describe the shape, center, and spread of each group.

```{r}
gerrymander <- gerrymander %>%
  mutate(margin16 = clinton16 - trump16)
```

```{r}
gerrymander %>%
  group_by(gerry) %>%
  summarise(
    min = min(margin16),
    mean = mean(margin16),
    max = max(margin16)
  )
```

```{r}
ggplot(gerrymander, aes(x = gerry, y = margin16)) +
  geom_violin() + 
  geom_boxplot(width = 0.1) +
  labs(
    x = "Level of gerrymandering",
    y = "Clinton win/loss margin (%)",
    title = "2016 Presidential margin by gerrymandering prevalence",
    subtitle = "Distribution of Clinton's margin over Trump's"
  )
```

* Given how similar the plots are across each level of gerrymandering, there doesn't seem to be a strong correlation between gerrymandering prevalence and how lopsided the 2016 presidential race was in that district. The means (centers) across each level of gerrymandering are similar (`low`: 6.32%, `mid`: 3.25%, `high`: 1.34%), and each plot has a similar shape and spread, with slight "right" (upward) skews and wide spreads all around 140% (`low`: -54.9% to 88.9%, `mid`: -63.3% to 82.8%, `high`: -63.1% to 78.2%).
